On the Gravitational stability of gravito-turbulent accretion disks

### 2016-02-02 paper/application.tex
to be submitted version done

Although our disk is global in extent, here we consider the local
stability at each radius. {\bf discuss: linear prob has no vrad in
  base state} 
%KMK I disagree with neglect of viscous flow in that there is a global Mdot. Revised version:
We are thus neglecting any global instabilities \citep{adams89,lodato05,kratter10} as well as any 
 evolution of the disk properties in response to GI.
% We are thus neglecting the slow, background viscous radial flow; as well
%as any global gravitational instabilities \citep{lodato05}. {\bf more refs?}

@@

Note that Eq. \ref{real_cool} falls within our definition of a beta cooling prescription, because
it is an explicit function of the thermodynamic states. However, we
formulated the linear problem with the standard beta cooling function 
given by Eq. \ref{beta_cool}, which has a different (less realistic) dependence on disk temperature.
 In order to adapt the existing  framework to the above PPD cooling function, we 
need to identify the equivalent $\beta$ and $\theta$ parameters that are required for the 
linearized equations (Eq. \ref{linear_beta}).

@@

%{\bf as opposed to comparing their unperturbed forms, as done in literature}
Linearizing Eq. \ref{real_cool} gives   
\begin{align}\label{linear_real_cool}
  (\gamma-1)\frac{\delta\Lambda}{\Sigma} = \frac{2\sigma(\gamma-1)
    T^4C_1}{f(\tau)c_{s}^2\Sigma}\left(\frac{\delta P}{\Sigma} -
  \frac{C_2}{C_1}c_{s}^2\frac{\delta\Sigma}{\Sigma}\right), 
\end{align}
where
\begin{align}
C_1(\tau, T) &= 4 - b\times g(\tau, T),\\ 
C_2(\tau, T) &= 4 + (1-b)\times g(\tau, T),\\ 
  g(\tau, T) &= \left( \frac{\tau^2-1}{\tau^2+1}\right)\left(1 -
  \frac{T_\mathrm{irr}^4}{T^4}\right). \label{g_def}
\end{align}
Comparing Eq. \ref{linear_real_cool} with the linearized form of the
standard cooling function, Eq. \ref{linear_beta}, we identify
\begin{align}
  &\beta = \frac{f(\tau)c_s^2\Sigma\Omega}{2\sigma(\gamma-1)C_1T^4},\label{real_beta}\\
  &\theta = \frac{C_2}{C_1},\label{real_theta} 
\end{align}
to be used in the 2D dispersion relation (Eq. \ref{thindisk}). 
Eq. \ref{real_beta} still represents a physical cooling 
time, and is consistent with previous definitions within factors of
order unity \citep[e.g.][their Eq. 2]{kratter10}.

%KMK
However, $\theta$ is no longer a direct proxy for the irradiation level
as it was in the standard beta cooling model.
%is now an effective irradiation level. 
Eq. \ref{real_theta} shows that $\theta$ is related to the true
irradiation temperature $\tirr$ through the function $g$ given by 
Eq. \ref{g_def}, and is therefore a only a weak function of the irradiation temperature. 
A physically non-irradiated disk with $\tirr=0$ will
have a non-zero value of the `irradiation parameter', $\theta\neq0$. More specifically $\theta = O(1)$ 
for all $\tirr <T$, and for our adopted opacity law, Eq. \ref{opacity_law}, 
\begin{align*}
  \theta = \frac{4-g}{4-2g}. 
\end{align*}
Thus for $T_\mathrm{irr} = 0$ we have $5/6<\theta<3/2$ by considering
$\tau\to 0,\,\infty$. However, for $T=\tirr$ we have $\theta = 1$, as
expected intuitively.    
%%KMK 
%This results from the fact that PPD cooling generally depends on both
%$P$ and $\Sigma$, but in the absence of irradiation standard beta
%cooling only depends on $P$. 
{\bf KMK: we should discuss}

@@

The fiducial disk is subject to viscous GI on dynamical timescales
($\lesssim 10$ orbits) for $R\gtrsim60$AU. We note this transition
radius is also implied by Eq. 16 in \cite{kratter10}. Coincidentally,
beyond this radius $\alpha\gtrsim 0.1$ and $\beta\lesssim 3$, which is
often quoted as criteria for disk fragmentation  
\citepalias[e.g.][]{rafikov15}. Thus viscous GI may be responsible for 
the transition between gravito-turbulence and fragmentation due to the
removal of rotational support by viscous (turbulent) stresses.

>>>>

Although our disk is global in extent, here we consider the local
stability at each radius. We are thus neglecting any global
instabilities \citep{adams89,lodato05,kratter10} as well as any
evolution of the disk properties in response to GI.

% We are thus neglecting the slow, background viscous radial flow; as well
%as any global gravitational instabilities \citep{lodato05}. {\bf more refs?}

@@

Note that Eq. \ref{real_cool} falls within our definition of a beta
cooling prescription, because 
it is an explicit function of the thermodynamic states. However, we
formulated the linear problem with the standard beta cooling function 
given by Eq. \ref{beta_cool}, which has a different (less realistic)
dependence on disk temperature. 
 In order to adapt the existing  framework to the above PPD cooling function, we 
need to identify the equivalent $\beta$ and $\theta$ parameters that are required for the 
linearized equations (see \S\ref{linear_bcool}).

@@

%{\bf as opposed to comparing their unperturbed forms, as done in literature}
Linearizing Eq. \ref{real_cool} gives   
\begin{align}\label{linear_real_cool}
  (\gamma-1)\frac{\delta\Lambda}{\Sigma} = \frac{2\sigma(\gamma-1)
    T^4C_1}{f(\tau)c_{s}^2\Sigma}\left(\frac{\delta P}{\Sigma} -
  \frac{C_2}{C_1}c_{s}^2\frac{\delta\Sigma}{\Sigma}\right), 
\end{align}
where
\begin{align}
C_1(\tau, T) &= 4 - b\times g(\tau, T),\\ 
C_2(\tau, T) &= 4 + (1-b)\times g(\tau, T),\\ 
  g(\tau, T) &= \left( \frac{\tau^2-1}{\tau^2+1}\right)\left(1 -
  \frac{T_\mathrm{irr}^4}{T^4}\right). \label{g_def}
\end{align}
Comparing Eq. \ref{linear_real_cool} with the linearized form of the
standard cooling function, Eq. \ref{linear_beta}, we identify
\begin{align}
  &\beta = \frac{f(\tau)c_s^2\Sigma\Omega}{2\sigma(\gamma-1)C_1T^4},\label{real_beta}\\
  &\theta = \frac{C_2}{C_1},\label{real_theta} 
\end{align}
to be used in the 2D dispersion relation (Eq. \ref{thindisk}). 
Eq. \ref{real_beta} represents a physical cooling \emph{for the
  perturbations}, and is consistent with previous definitions within
factors of order unity \citep[e.g.][their Eq. 2]{kratter10}.

%KMK
%However, $\theta$ is no longer a direct proxy for the irradiation level
%as it was in the standard beta cooling model.
%is now an effective irradiation level. 
Eq. \ref{real_theta} shows that the re-defined $\theta$ is related to
the true irradiation temperature $\tirr$ through the function $g$ given by 
Eq. \ref{g_def}, and is therefore a only a weak function of the irradiation temperature. 
%A physically non-irradiated disk with $\tirr=0$ will 
%have a non-zero value of the `irradiation parameter', $\theta\neq0$. 
More specifically $\theta = O(1)$ 
for all $\tirr <T$, and for our adopted opacity law, Eq. \ref{opacity_law}, 
\begin{align*}
  \theta = \frac{4-g}{4-2g}. 
\end{align*}
Thus for $T_\mathrm{irr} = 0$ we have $5/6<\theta<3/2$ by considering
$\tau\to 0,\,\infty$. However, for $T=\tirr$ we have $\theta = 1$, as
expected intuitively.    
%%KMK 
%This results from the fact that PPD cooling generally depends on both
%$P$ and $\Sigma$, but in the absence of irradiation standard beta
%cooling only depends on $P$. 
%{\bf KMK: we should discuss}

@@

The fiducial disk is subject to viscous GI on dynamical timescales
($\lesssim 10$ orbits) for $R\gtrsim60$AU. We note this transition
radius is also implied by Eq. 16 in \cite{kratter10}. Coincidentally,
beyond this radius $\alpha\gtrsim 0.1$ (and 
$\beta\lesssim 3$), which is often quoted as a condition for disk
fragmentation   
\citepalias[e.g.][]{rafikov15}. Thus viscous GI may be responsible for 
the transition between gravito-turbulence and fragmentation due to the
removal of rotational support by viscous (turbulent) stresses.

### 2016-02-02 paper/linear3d.tex
revision 3 by MKL
to be submitted version done

The linearized continuity, momentum and energy equations are
\begin{align}
   &s \delta \rho = - \ii k \rho\delta v_x -
  \left(\rho\delta v_z\right)^\prime \label{lin_mass} \\ 
   &s  \delta v_x = - \ii k\frac{\delta P}{\rho} - \ii k \delta 
  \Phi + 2\Omega \delta v_y + \delta F_x,\\
   &s \delta v_y = (q-2)\Omega\delta v_x + \delta F_y,\\
%   &s \delta v_z = - \left(\frac{\delta P}{\rho}\right)^\prime -
%  \delta\Phi^\prime +
%  \left(\ln{\rho}\right)^\prime\left(c_s^2\frac{\delta\rho}{\rho} -
%  \frac{\delta P}{\rho}\right) + \delta F_z,\\
  &s \delta v_z =
  \left(\ln{P}\right)^\prime c_s^2\frac{\delta\rho}{\rho} -
  \left(\frac{\delta P}{\rho}\right)^\prime -
  \left(\ln{\rho}\right)^\prime\frac{\delta P}{\rho} + \delta F_z, \\ 
  &s \frac{\delta P}{\rho} = 
%- \ii \gamma k \frac{c_s^2}{\Gamma}\delta v_x 
 - \ii k \gamma c_s^2\delta v_x
%- c_s^2\left[
%  \left(\ln{\rho}\right)^\prime\delta v_z + 
%  \frac{\gamma}{\Gamma}\delta v_z^\prime\right] \notag\\
- c_s^2\left[
 \left(\ln{P}\right)^\prime\delta v_z + 
  \gamma\delta v_z^\prime\right] \notag\\
   & \phantom{s \frac{\delta P}{\rho}=}
  + (\gamma-1)\frac{\delta\mathcal{H}_\mathrm{visc}}{\rho} 
  -(\gamma-1)\frac{\delta\Lambda}{\rho}
%  - \frac{1}{t_c}\left(\frac{\delta P}{\rho} -
%  \frac{\theta}{\Gamma}c_s^2\frac{\delta\rho}{\rho}\right) 
  \label{lin_energy},\\
  &\delta\Phi^{\prime\prime} - k^2\delta\Phi =
  \frac{\Omega^2}{Q_\mathrm{3D}}\left(\frac{\delta\rho}{\rho_0}\right)\label{lin_gravity},  
\end{align}
where $^\prime$ denotes $d/dz$. The perturbed viscous forces are {\bf F is also radiative flux....}
\begin{align}
  \delta F_x =& \nu \left[\delta v_x^{\prime\prime} + 
    \left(\ln{\rho\nu}\right)^\prime \delta v_x^\prime - 
     \frac{4}{3}k^2\delta v_x \right] - \nu_bk^2\delta v_x \notag\\
   &+ \ii \nu k \left[\frac{1}{3}\delta v_z^\prime+ 
    \left(\ln{\rho\nu}\right)^\prime\delta v_z\right] + \ii k \nu_b
  \delta v_z^\prime,\\
  \delta F_y = & \nu  \left[\delta v_y^{\prime\prime} + 
    \left(\ln{\rho\nu}\right)^\prime \delta v_y^\prime - 
     k^2\delta v_y \right] - \ii\nu kq\Omega\delta\ln{\rho\nu},\\
  \delta F_z =& \nu\left[\frac{4}{3}\delta v_z^{\prime\prime} + 
    \frac{4}{3}\left(\ln{\rho\nu}\right)^\prime \delta v_z^\prime - 
    k^2\delta v_z \right] \notag\\
  & +\nu_b\left[\delta v_z^{\prime\prime} + \left(\ln{\rho\nu_b}\right)^\prime\delta v_z^\prime\right]\notag\\
  & + \ii \nu k  \left[\frac{1}{3}\delta v_x^\prime -  
    \frac{2}{3}\left(\ln{\rho\nu}\right)^\prime\delta v_x\right]\notag\\
  &+\ii \nu_b k \left[\delta v_x^\prime + \left(\ln{\rho\nu_b}\right)^\prime\delta v_x\right],
%  &\delta f_x = -\frac{4}{3} \nu k^2\Sigma \delta v_x,\\
%  &\delta f_y = -\ii q k \Omega \delta(\nu\Sigma) - \nu k^2
%  \Sigma\delta v_y,\\
%  &\delta Q_\mathrm{visc} = (q\Omega)^2\delta(\nu\Sigma) - 2 \ii \nu 
%  k q\Omega\Sigma\delta v_y,\\
%  &\delta(\nu\Sigma) =
%  \frac{\alpha}{\Omega}\left[c_s^2(1+\mu)\delta\Sigma + \lambda\delta P\right].
%  &\delta f_y = - \ii \nu q k \Omega \delta \Sigma - \nu k^2\Sigma\delta
%  v_y,\\
%  &\delta Q_\mathrm{visc} = \nu
%  \left(q\Omega\right)^2\delta\Sigma - 2\ii\nu q k \Omega\Sigma\delta
%  v_y. 
\end{align}
and the perturbed viscous heating is given by 
\begin{align}
  \frac{\delta\mathcal{H}_\mathrm{visc}}{\rho}=& \nu (q\Omega)^2
  \delta\ln{\rho\nu} - 2\ii\nu k q \Omega \delta v_y,\\
  \delta\ln{\rho\nu} =& (1+\mu)\frac{\delta\rho}{\rho} +
  \frac{\lambda}{c_s^2}\frac{\delta P}{\rho}. \label{linear_beta_cool}
\end{align}
In Eq. \ref{lin_gravity}, the 3D self-gravity parameter is
\begin{align}\label{Q3d_def}
  Q_\mathrm{3D} \equiv \frac{\Omega^2}{4\pi G \rho_0}. 
\end{align}
The linearized cooling functions $\delta\Lambda$ are given below. 
Eq. \ref{lin_mass}---\ref{lin_gravity}, supplemented with appropriate
boundary conditions, constitutes an eigenvalue problem for the growth
rate $s$.

\subsection{Linearized beta cooling}
For the standard beta cooling prescription, linearizing
Eq. \ref{beta_cool} gives

@@

\subsection{Linearized radiative cooling}
Linearizing Eq. \ref{rad_cool1}---\ref{rad_cool2} with the
temperature-dependent opacity law in Eq. \ref{opacity_law} gives
\begin{align}
  \frac{\delta \Lambda}{\rho} 
  =& \frac{16\sigma T^3}{3\kappa_d\rho^2} k^2 \delta T \notag\\
  & - \frac{16\sigma }{3\rho}\frac{d}{dz}\left\{
  \frac{T^3}{\kappa_d\rho}\left[\delta T^\prime +
    (3-b)\left(\ln{T}\right)^\prime \delta T - T^\prime \frac{\delta\rho}{\rho}\right]
  \right\}.\label{linear_rad_cool}
\end{align}
Since Eq. \ref{linear_rad_cool} contains vertical derivatives of the
perturbations, it is not generically possible to map radiative cooling 
to the beta cooling prescription, except for special problems
\citep[e.g.][]{lin15}. %KMK
We now consider the behavior of two and three dimensional disks in the
non-ideal physics, linear GI framework {\bf or some other transition sentence}.

>>>>

The linearized continuity, momentum and energy equations are
\begin{align}
   &s \delta \rho = - \ii k \rho\delta v_x -
  \left(\rho\delta v_z\right)^\prime \label{lin_mass} \\ 
   &s  \delta v_x = - \ii k\frac{\delta P}{\rho} - \ii k \delta 
  \Phi + 2\Omega \delta v_y + \delta F_x,\\
   &s \delta v_y = (q-2)\Omega\delta v_x + \delta F_y,\\
%   &s \delta v_z = - \left(\frac{\delta P}{\rho}\right)^\prime -
%  \delta\Phi^\prime +
%  \left(\ln{\rho}\right)^\prime\left(c_s^2\frac{\delta\rho}{\rho} -
%  \frac{\delta P}{\rho}\right) + \delta F_z,\\
  &s \delta v_z =
  \left(\ln{P}\right)^\prime c_s^2\frac{\delta\rho}{\rho} -
  \left(\frac{\delta P}{\rho}\right)^\prime -
  \left(\ln{\rho}\right)^\prime\frac{\delta P}{\rho} + \delta F_z, \\ 
  &s \frac{\delta P}{\rho} = 
%- \ii \gamma k \frac{c_s^2}{\Gamma}\delta v_x 
 - \ii k \gamma c_s^2\delta v_x
%- c_s^2\left[
%  \left(\ln{\rho}\right)^\prime\delta v_z + 
%  \frac{\gamma}{\Gamma}\delta v_z^\prime\right] \notag\\
- c_s^2\left[
 \left(\ln{P}\right)^\prime\delta v_z + 
  \gamma\delta v_z^\prime\right] \notag\\
   & \phantom{s \frac{\delta P}{\rho}=}
  + (\gamma-1)\frac{\delta\mathcal{H}_\mathrm{visc}}{\rho} 
  -(\gamma-1)\frac{\delta\Lambda}{\rho}
%  - \frac{1}{t_c}\left(\frac{\delta P}{\rho} -
%  \frac{\theta}{\Gamma}c_s^2\frac{\delta\rho}{\rho}\right) 
  \label{lin_energy},\\
  &\delta\Phi^{\prime\prime} - k^2\delta\Phi =
  \frac{\Omega^2}{Q_\mathrm{3D}}\left(\frac{\delta\rho}{\rho_0}\right)\label{lin_gravity},  
\end{align}
where $^\prime$ denotes $d/dz$. The perturbed viscous forces are %{\bf F is also radiative flux....}
\begin{align}
  \delta F_x =& \nu \left[\delta v_x^{\prime\prime} + 
    \left(\ln{\rho\nu}\right)^\prime \delta v_x^\prime - 
     \frac{4}{3}k^2\delta v_x \right] - \nu_bk^2\delta v_x \notag\\
   &+ \ii \nu k \left[\frac{1}{3}\delta v_z^\prime+ 
    \left(\ln{\rho\nu}\right)^\prime\delta v_z\right] + \ii k \nu_b
  \delta v_z^\prime,\\
  \delta F_y = & \nu  \left[\delta v_y^{\prime\prime} + 
    \left(\ln{\rho\nu}\right)^\prime \delta v_y^\prime - 
     k^2\delta v_y \right] - \ii\nu kq\Omega\delta\ln{\rho\nu},\\
  \delta F_z =& \nu\left[\frac{4}{3}\delta v_z^{\prime\prime} + 
    \frac{4}{3}\left(\ln{\rho\nu}\right)^\prime \delta v_z^\prime - 
    k^2\delta v_z \right] \notag\\
  & +\nu_b\left[\delta v_z^{\prime\prime} + \left(\ln{\rho\nu_b}\right)^\prime\delta v_z^\prime\right]\notag\\
  & + \ii \nu k  \left[\frac{1}{3}\delta v_x^\prime -  
    \frac{2}{3}\left(\ln{\rho\nu}\right)^\prime\delta v_x\right]\notag\\
  &+\ii \nu_b k \left[\delta v_x^\prime + \left(\ln{\rho\nu_b}\right)^\prime\delta v_x\right],
%  &\delta f_x = -\frac{4}{3} \nu k^2\Sigma \delta v_x,\\
%  &\delta f_y = -\ii q k \Omega \delta(\nu\Sigma) - \nu k^2
%  \Sigma\delta v_y,\\
%  &\delta Q_\mathrm{visc} = (q\Omega)^2\delta(\nu\Sigma) - 2 \ii \nu 
%  k q\Omega\Sigma\delta v_y,\\
%  &\delta(\nu\Sigma) =
%  \frac{\alpha}{\Omega}\left[c_s^2(1+\mu)\delta\Sigma + \lambda\delta P\right].
%  &\delta f_y = - \ii \nu q k \Omega \delta \Sigma - \nu k^2\Sigma\delta
%  v_y,\\
%  &\delta Q_\mathrm{visc} = \nu
%  \left(q\Omega\right)^2\delta\Sigma - 2\ii\nu q k \Omega\Sigma\delta
%  v_y. 
\end{align}
and the perturbed viscous heating is given by 
\begin{align}
  \frac{\delta\mathcal{H}_\mathrm{visc}}{\rho}=& \nu (q\Omega)^2
  \delta\ln{\rho\nu} - 2\ii\nu k q \Omega \delta v_y,\\
  \delta\ln{\rho\nu} =& (1+\mu)\frac{\delta\rho}{\rho} +
  \frac{\lambda}{c_s^2}\frac{\delta P}{\rho}. \label{linear_beta_cool}
\end{align}
In Eq. \ref{lin_gravity}, the 3D self-gravity parameter is
\begin{align}\label{Q3d_def}
  Q_\mathrm{3D} \equiv \frac{\Omega^2}{4\pi G \rho_0}. 
\end{align}
The linearized cooling functions $\delta\Lambda$ are given below. 
Eq. \ref{lin_mass}---\ref{lin_gravity}, supplemented with appropriate
boundary conditions, constitutes an eigenvalue problem for the growth
rate $s$.

\subsection{Linearized beta cooling}\label{linear_bcool}
For the standard beta cooling prescription, linearizing
Eq. \ref{beta_cool} gives

@@

Note that any linearized beta cooling function can be put into the
form of Eq. \ref{linear_beta} with appropriate definitions of $t_c$
and $\theta$ (see \S\ref{ppd_cooling} for an example).    
For the stability problem we may simply regard $\theta$ as a parameter 
for the density-dependence of any generic beta cooling function. If 
we specifically consider standard beta cooling, then $\theta$ also
represents physical irradiation.

\subsection{Linearized radiative cooling}
Linearizing Eq. \ref{rad_cool1}---\ref{rad_cool2} with the
temperature-dependent opacity law in Eq. \ref{opacity_law} gives
\begin{align}
  \frac{\delta \Lambda}{\rho} 
  =& \frac{16\sigma T^3}{3\kappa_d\rho^2} k^2 \delta T \notag\\
  & - \frac{16\sigma }{3\rho}\frac{d}{dz}\left\{
  \frac{T^3}{\kappa_d\rho}\left[\delta T^\prime +
    (3-b)\left(\ln{T}\right)^\prime \delta T - T^\prime \frac{\delta\rho}{\rho}\right]
  \right\}.\label{linear_rad_cool}
\end{align}
Since Eq. \ref{linear_rad_cool} contains vertical derivatives of the
perturbations, it is not generically possible to map radiative cooling 
to the beta cooling prescription, except for special problems
\citep[e.g.][]{lin15}.

We now consider the gravitational stability of two and three 
dimensional disks in the presence of non-ideal
physics: cooling and viscosity.  
%{\bf or some other transition sentence}.

### 2016-02-02 paper/model3d.tex
to be submitted version done

\subsubsection{$\beta$ cooling}\label{beta_cool_model}
In our $\beta$ cooling model, the energy loss per unit volume is specified  
as an explicit function of the thermodynamic variables. 
A prototypical example is 
\begin{align}\label{beta_cool}
  \Lambda(\rho, T) =
  %  \frac{P}{(\gamma-1)T}\frac{\left(T-T_\mathrm{irr}\right)}{t_c}, 
  \frac{\mathcal{R}\rho}{(\gamma-1)}\frac{\left(T-T_\mathrm{irr}\right)}{t_c}. 
\end{align}
Recall  $\tcool = \beta\Omega^{-1}$ is 
the cooling timescale with $\beta$ a constant input parameter. Here, 
\begin{align}\label{tirr_def}
  \tirr = \theta T_\mathrm{eq}
\end{align}
is a reference temperature field parameterized as a constant 
fraction $\theta$ of the equilibrium temperature
$T_\mathrm{eq}$. Physically, $\tirr$ is the floor temperature set by,
for example, stellar or background irradiation. We will thus refer to
$\theta$ as the `irradiation level'. {\bf ASK} As we discuss in
\S\ref{2d_inviscid}, the limits $\theta \rightarrow 0, \theta
\rightarrow 1$ do not exactly correspond to the limits of
non-irradiated or irradiation-dominated disks as discussed in the
literature \citep{kratter11,rice11}. 
%stellar illum or envelope rad

@@

Note that, although we assume standard beta cooling in formulating the
linear problem later, any explicit 
cooling function, say $\Lambda_1(\rho,T)$, can be mapped to
that obtained from Eq. \ref{beta_cool} by equating their linearized
forms, i.e. $\delta \Lambda \equiv \delta \Lambda_1$. This then 
defines the $\beta$ and $\theta$ parameters to be used in the 
framework we develop. We do this in \S\ref{2dppd} where we adopt a
more realistic beta cooling function for PPDs. In that case, $\theta$
may or may not represent a physical irradiation.

>>>>

\subsubsection{$\beta$ cooling}\label{beta_cool_model}
In our $\beta$ cooling model, the energy loss per unit volume is specified  
as an explicit function of the thermodynamic variables. 
A prototypical example is 
\begin{align}\label{beta_cool}
  \Lambda(\rho, T) =
  %  \frac{P}{(\gamma-1)T}\frac{\left(T-T_\mathrm{irr}\right)}{t_c}, 
  \frac{\mathcal{R}\rho}{(\gamma-1)}\frac{\left(T-T_\mathrm{irr}\right)}{t_c}. 
\end{align}
Recall  $\tcool = \beta\Omega^{-1}$ is 
the cooling timescale with $\beta$ a constant input parameter. Here, 
\begin{align}\label{tirr_def}
  \tirr = \theta T_\mathrm{eq}
\end{align}
is a reference temperature field parameterized as a constant 
fraction $\theta$ of the equilibrium temperature
$T_\mathrm{eq}$. Physically, $\tirr$ is the floor temperature set by,
for example, stellar or background irradiation. Note that $\theta$  
should only be interpreted as an irradiation parameter when it is 
defined through Eq. \ref{tirr_def} and 
Eq. \ref{beta_cool} adopted as the cooling function.

@@

%We shall need to re-define $\theta$ and $\beta$  
%when considering other cooling functions (see below).

We assume standard beta cooling in formulating the linear
problem. However, the corresponding linear problem for  
any other explicit cooling function, say $\Lambda_1(\rho,T)$, can be  
obtained by equating its linearized form to 
that of Eq. \ref{beta_cool}, i.e. setting $\delta \Lambda \equiv
\delta \Lambda_1$. This then defines the $\beta$ and $\theta$
parameters to be used in the framework we develop later (see also 
\S\ref{linear_bcool}). We do this in \S\ref{2dppd} where we adopt a  
more realistic beta cooling function for PPDs. In that case, $\theta$ 
may or may not directly represent a physical irradiation.

### 2016-02-02 paper/paper3d.tex
to be submitted version done

\begin{abstract}
  Low mass, self-gravitating accretion disks admit quasi-steady,
  `gravito-turbulent' states in which cooling balances turbulent viscous 
  heating. However, numerical simulations show that gravito-turbulence
  cannot be sustained beyond dynamical timescales when the cooling
  rate or corresponding turbulent viscosity is too large. The result is disk fragmentation.         
  We motivate and quantify an interpretation of disk 
  fragmentation as the inability to maintain gravito-turbulence due to  
  formal secondary instabilities driven by:  
\begin{inparaenum}[1)] 
\item 
  cooling, which reduces pressure support; and/or
\item 
  viscosity, which reduces rotational support. 
\end{inparaenum}
%We analyze the gravitational stability of accretion disks 
%including these non-ideal effects explicitly.    
%We derive a general dispersion relation for linear waves in a 
%self-gravitating, viscous, non-adiabatic disk with internal heating,
%external irradiation, and cooling. 
We analyze the gravitational stability of viscous, non-adiabatic
accretion disks with internal heating, external irradiation, and cooling.  
We consider parameterized cooling functions in two and three dimensions,
as well as 
radiative diffusion in 3D. We show that 
generally there is no critical cooling rate/viscosity below which 
the disk is formally stable, although interesting limits appear 
for unstable modes with lengthscales on the order of the disk thickness.   
%when we examine the growth 
%of modes of order the disk scale height. 
We apply this new linear theory to 
global models of protoplanetary disks (PPDs) subject to
gravito-turbulence modeled as an effective 
viscosity, and cooling via dust opacity. 
We find that viscosity renders the PPD beyond $\sim 60$AU dynamically
unstable on radial %this is for Mdot = 1d-6  
lengthscales a few times the local disk thickness. This is coincident 
with the empirical condition for disk fragmentation based on a
maximum sustainable stress. 
We suggest turbulent stresses can play an active role in realistic disk 
fragmentation by removing rotational stabilization against 
self-gravity, and that the observed transition in behavior from 
gravito-turbulent to fragmenting may derive from  
instability of the gravito-turbulent state itself.
\end{abstract}

@@

%%KMK 
Previous work has quantified the role of cooling primarily as a means to reduce the sound speed term
in $Q$ \citep[though see e.g.,][]{clarke07}. Here we will quantify how cooling enables GI
even when $Q>1$. For example, if perturbations can
cool to zero temperature, we find the above dispersion relation is
modified to read 
\begin{align}\label{cool_dr}
s^2 = 2\pi G\Sigma |k| - \kappa^2 - \gamma \left(\frac{\tcool s}{1 +
  \tcool s}\right)c_s^2k^2,
\end{align}
where $\gamma$ is the adiabatic index. Cooling increases the growth
rate by reducing the magnitude of the 
pressure term. In fact, for $t_c\in[0,\infty)$ a formal condition
for instability is  
\begin{align*}
  2\pi G \Sigma |k| > \kappa^2,
\end{align*}
which is what would be obtained from Eq. \ref{classic_dr} with 
pressure neglected ($c_s\to 0$). This condition does not actually
depend on the cooling time, and instability is
possible for any finite $Q$. %at sufficiently small scales 
Thus, the mere presence of cooling changes the qualitative nature of
GI compared to the simple Toomre condition. 
%singular limit

@@

For an isothermal, viscous, self-gravitating disk in 2D,
\cite{gammie96} finds the approximate dispersion relation   
\begin{align}
s \simeq \frac{\nu k^2\left(2\pi G\Sigma |k| -
  c_s^2k^2\right)}{\kappa^2 + c_s^2k^2 - 2\pi G\Sigma|k|}. 
\end{align}
Assuming $Q>1$, instability occurs if 
\begin{align*}
  2\pi G \Sigma > c_s^2|k|,
\end{align*}
provided that $\nu k^2\neq0$. This condition for viscous GI is 
identical to what would be obtained from Eq. \ref{classic_dr} with
rotation neglected ($\Omega\to 0$). That is, a classically stable
disk can be destabilized by viscosity as it effectively eliminates
rotational stabilization \citep{lynden-bell74}.

@@

\subsection{Relevance to gravito-turbulent disk fragmentation}\label{frag_intro}
We develop a general framework for viscous disks without  
assuming a specific origin of the turbulence that this viscosity 
represents. 
We will, however, apply the theory to `gravito-turbulent'
disks in which the viscosity is associated with some underlying
(classic) GI, as described below.

>>>>

\begin{abstract}
  Low mass, self-gravitating accretion disks admit quasi-steady,
  `gravito-turbulent' states in which cooling balances turbulent viscous 
  heating. However, numerical simulations show that gravito-turbulence
  cannot be sustained beyond dynamical timescales when the cooling
  rate or corresponding turbulent viscosity is too large. The result is disk fragmentation.         
  We motivate and quantify an interpretation of disk 
  fragmentation as the inability to maintain gravito-turbulence due to  
  formal secondary instabilities driven by:  
\begin{inparaenum}[1)] 
\item 
  cooling, which reduces pressure support; and/or
\item 
  viscosity, which reduces rotational support. 
\end{inparaenum}
%We analyze the gravitational stability of accretion disks 
%including these non-ideal effects explicitly.    
%We derive a general dispersion relation for linear waves in a 
%self-gravitating, viscous, non-adiabatic disk with internal heating,
%external irradiation, and cooling. 
We analyze the gravitational stability of viscous, non-adiabatic
accretion disks with internal heating, external irradiation, and cooling.  
We consider parameterized cooling functions in two and three dimensions,
as well as 
radiative diffusion in 3D. We show that 
generally there is no critical cooling rate/viscosity below which 
the disk is formally stable, although interesting limits appear 
for unstable modes with lengthscales on the order of the disk thickness.   
%when we examine the growth 
%of modes of order the disk scale height. 
We apply this new linear theory to 
global models of protoplanetary disks subject to
gravito-turbulence modeled as an effective 
viscosity, and cooling via dust opacity. 
We find that viscosity renders the disk beyond $\sim 60$AU dynamically
unstable on radial %this is for Mdot = 1d-6  
lengthscales a few times the local disk thickness. This is coincident 
with the empirical condition for disk fragmentation based on a
maximum sustainable stress. 
We suggest turbulent stresses can play an active role in realistic disk 
fragmentation by removing rotational stabilization against 
self-gravity, and that the observed transition in behavior from 
gravito-turbulent to fragmenting may derive from  
instability of the gravito-turbulent state itself.
\end{abstract}

@@

%%KMK 
Previous work has quantified the role of cooling primarily as a means to reduce the sound speed term
in $Q$ \citep[though see e.g.,][]{clarke07}. Here we will quantify how cooling enables GI
even when $Q>1$. For example, if perturbations can
cool to arbitrarily low temperatures (which, in fact, is a common
cooling prescription in numerical simulations), we find the above
dispersion relation is  
modified to read 
\begin{align}\label{cool_dr}
s^2 = 2\pi G\Sigma |k| - \kappa^2 - \gamma \left(\frac{\tcool s}{1 +
  \tcool s}\right)c_s^2k^2,
\end{align}
where $\gamma$ is the adiabatic index. Cooling increases the growth
rate by reducing the magnitude of the 
pressure term. In fact, for $t_c\in[0,\infty)$ a formal condition
for instability is  
\begin{align*}
  2\pi G \Sigma |k| > \kappa^2,
\end{align*}
which is what would be obtained from Eq. \ref{classic_dr} with 
pressure neglected ($c_s\to 0$). This condition does not actually
depend on the cooling time, and instability is
possible for any finite $Q$. %at sufficiently small scales 
Thus, the mere presence of cooling changes the qualitative nature of
GI compared to the simple Toomre condition. 
%singular limit

@@

For an isothermal, viscous, self-gravitating disk in 2D,
\cite{gammie96} finds the approximate dispersion relation   
\begin{align}
s \simeq \frac{\nu k^2\left(2\pi G\Sigma |k| -
  c_s^2k^2\right)}{\kappa^2 + c_s^2k^2 - 2\pi G\Sigma|k|}. 
\end{align}
Assuming $Q>1$, instability occurs if 
\begin{align*}
  2\pi G \Sigma > c_s^2|k|,
\end{align*}
provided that $\nu k^2\neq0$. This condition for viscous GI is 
identical to what would be obtained from Eq. \ref{classic_dr} with
rotation neglected ($\Omega\to 0$). That is, a classically stable
disk can be destabilized by viscosity as it reduces 
rotational stabilization \citep{lynden-bell74}.

@@

\subsection{Relevance to gravito-turbulent disk fragmentation}\label{frag_intro}
We develop a general framework for viscous disks without  
assuming a specific origin for the viscosity. 
We will, however, apply the theory to `gravito-turbulent'
disks in which the viscosity is associated with some underlying
(classic) GI, as described below.
